Improved guarantees and a multiple-descent curve for \\ Column Subset Selection and the Nystr\"om method

### 2022-08-23 design/talk/Michal_Derezinski_Statement_new.tex
Commit original repos

\begin{document}
\maketitle
%\section{}
%\subsection{}

A key motivation for our work is interpretable data summarization,
which is a core challenge in reasoning 
about large datasets and machine learning models.  Data summarization
can be used, for example, to select a representative subset of gene
variants from a genetics dataset, or a collection of most informative
documents from a text database.  When data are represented
numerically, they are often described via matrices, in which case
linear algebra suggests a natural (and in a certain sense optimal) way
of performing data summarization: find the principal components
corresponding to the largest directions of variance.  These principal
components work well for black-box models that are evaluated only in
terms of prediction quality, but they are generally not interpretable
in terms of the domain from which the data are drawn. They do not
correspond to, say, a particular document or a gene variant, but
rather a complex mixture of them.  Neural networks, on which the
machine learning community increasingly relies, simply exacerbate this
problem.  A long-standing challenge has been to find summaries of data
which mimic the numerical properties of principal components and which
are also interpretable.

The cost of interpretability, when formulated this way, has strong
connections to the so-called Column Subset Selection Problem in
Randomized Numerical Linear Algebra (RandNLA). It
has been extensively studied in the literature, resulting in optimal
worst-case guarantees, developed for the first time nearly fifteen
years ago. These results are still relied upon as an important
technical tool in recent works, including the best paper at last
year’s International Conference on Machine Learning, which applied
them to Gaussian Process regression. The worst-case results suggest
that the cost of interpretability should become progressively higher
as the data summaries get larger.
To verify this, we carefully constructed a worst-case example for this problem; and we
showed that, for this example, the cost of interpretability does reach
the worse-case bound for one size of the summary, but then it rapidly
drops down for both larger and smaller sizes. As it turns out, in this and other
worst-case examples, the high cost of interpretability is only a
corner case.

@@

To explain this, we go beyond worst-case analysis, and our results are
able to accurately capture the non-linear behavior of the cost of
interpretability. Specifically, we conclude that, except for certain
pathological examples which we can characterize, the cost of
interpretability is far smaller than suggested by prior work, and it
is often negligible for real-world problems. To construct our
interpretable summaries, we use a randomized subset selection
technique based on Determinantal Point Processes (DPPs), which provide
%technique called Determinantal Point Processes, which is 
a probabilistic model of diversity that has emerged across many scientific domains.
DPPs were first discovered in physics as a model of fermions in thermal
equilibrium, and they have since been used in random matrix theory, graph
theory and quantum mechanics. More recently, thanks to the emergence
of efficient algorithms for DPP sampling, they have also become
popular in RandNLA and Machine Learning.

@@

Both the double descent curve in machine learning, and the new
multiple-descent curve in data summarization, shown in our work, are
related to fundamental phase transitions observed in the behavior of
high-dimensional random matrices. Obtaining precise characterizations
of these phenomena is of crucial importance to understanding modern
machine learning as well as algorithmic-statistical tradeoffs that are
central to the foundations of data science.
\vspace{2mm}

>>>>

\begin{document}
\maketitle
%\section{}
%\subsection{}
\vspace{-5mm}

Thank you \textbf{very} much. We are \textbf{really} excited to receive this award. My
name is Micha{\l}~Derezi\'nski, I am a postdoctoral researcher at UC Berkeley.

A key motivation for our work is interpretable data summarization,
which is a core challenge in reasoning 
about large datasets and machine learning models.  Data summarization
can be used, for example, to select a representative subset of gene
variants from a genetics dataset, or a collection of most informative
documents from a text database.  When data are represented
numerically, they are often described via matrices, in which case
linear algebra suggests a natural (and in a certain sense optimal) way
of performing data summarization: namely, finding the principal components
corresponding to the largest directions of variance.  These principal
components work well for black-box models that are evaluated only in
terms of prediction quality, but they are generally not interpretable
in terms of the domain from which the data are drawn. They do not
correspond to, say, a particular document or a gene variant, but
rather a complex mixture of them.  Neural networks, on which the
machine learning community increasingly relies, simply exacerbate this
problem of non-interpretability.  A long-standing challenge has been to find summaries of data
which mimic the numerical properties of principal components and which
are also interpretable.

The cost of interpretability, when formulated this way, has strong
connections to the so-called Column Subset Selection Problem in
Randomized Numerical Linear Algebra. It
has been extensively studied in the literature, resulting in optimal
worst-case guarantees, developed for the first time nearly fifteen
years ago. These results are still relied upon as an important
technical tool in recent works, including the best paper at last
year’s International Conference on Machine Learning, which applied
them to Gaussian Process regression. The worst-case results suggest
that the cost of interpretability should become progressively higher
as the data summaries get larger.
To verify this, we carefully constructed a worst-case example for this problem; and we
showed that, for this example, the cost of interpretability \textbf{does} reach
the worse-case bound for \textbf{one} size of the summary, but then it rapidly
drops down for both larger and smaller sizes. As it turns out, in this and other
worst-case examples, the high cost of interpretability is only a
corner case.

@@

To explain this, we go beyond worst-case analysis, and our results are
able to accurately capture the non-linear behavior of the cost of
interpretability. Specifically, we conclude that, except for certain
pathological examples which we can characterize, the cost of
interpretability is far smaller than suggested by prior work, and it
is often negligible for real-world problems. To construct our
interpretable summaries, we use a randomized subset selection
technique based on Determinantal Point Processes (DPPs for short), which provide
%technique called Determinantal Point Processes, which is 
a probabilistic model of diversity that has emerged across many scientific domains.
DPPs were first discovered in physics as a model of fermions in thermal
equilibrium, and they have since been used in random matrix theory, graph
theory and quantum mechanics. More recently, thanks to the emergence
of efficient algorithms for DPP sampling, they have also become
popular in Randomized Numerical Linear Algebra and Machine~Learning.

@@

Both the double descent curve in machine learning, and the new
multiple-descent curve in data summarization, shown in our work, are
related to fundamental phase transitions observed in the behavior of
high-dimensional random matrices. Obtaining precise characterizations
of these phenomena is of crucial importance to understanding modern
machine learning as well as algorithmic-statistical tradeoffs that are
central to the foundations of data science.
\vfill
\textbf{Summary}:
Interpretable data summarization is a core challenge in reasoning
about large datasets and machine learning models.  Data summarization
can be used, e.g., to select a representative subset of gene variants
from a genetics dataset, or a collection of most informative documents
from a text database.  When data are represented numerically, they are
often described via matrices, in which case linear algebra suggests a
natural (and in certain ways optimal) way of performing data
summarization: find the principal components corresponding to the
largest directions of variance.  These principal components work well
for black-box models that are evaluated only in terms of prediction
quality, but they are generally not interpretable in terms of the
domain from which the data are drawn. They do not correspond to, say,
a particular document or a gene variant, but rather a complex mixture
of them.  Neural networks, on which the machine learning community
increasingly relies, simply exacerbate this problem.  A long-standing
challenge has been to find summaries of data which mimic the numerical
properties of principal components and which are also
interpretable. The added cost of this interpretability can be
significant, as shown by prior worst-case analysis.  In our work, we
exploit the inherent structure of data to show that, except for
pathological examples which we characterize, the cost of
interpretability is far smaller than suggested by prior worst-case
analysis, and it is often negligible for real-world problems.  Our
analysis reveals an intriguing new phenomenon, which we call the
multiple-descent curve in subset selection.  This is a phase
transition, observed through a spike in the cost of interpretability,
which occurs when the data exhibit an underlying hierarchical
structure.  The multiple-descent phenomenon can be observed in
practice, but it can also be avoided by tuning model parameters.

\newpage
\begin{itemize}
\item Interpretability is important, but Modern ML models are
  black boxes.
  \begin{quote}
    Obtaining interpretable data summaries is an
    important challenge, especially in the context of modern ML and
    Deep Learning models, which are often black boxes.
  \end{quote}
  \begin{quote}
    Our work suggests that obtaining interpretable summaries can be
    done effectively, if we ignore pathological problem instances,
    which do not occur in the real-world.
  \end{quote}
\item Going beyond worst-case analysis to understand different phases
  of learning.
  \begin{quote}
Theoretical work in ML and Data Science, particularly involving
techniques from Randomized Numerical Linear Algebra, often focuses on
the worst-case examples.
\end{quote}
\begin{quote}
  Worst-case analysis does not capture the
fundamental phase transitions associated with learning, such as
double-descent and multiple-descent.
\end{quote}
\begin{quote}
Using techniques from statistical physics and random matrix theory,
such as Determinantal Point Processes, we can go beyond the worst-case
analysis and obtain precise characterizations of these phenomena.
\end{quote}
\item The paper that “forced Timnit Gebru out of Google”
  \begin{quote}
    I don’t know the specifics of the situation with Timnit, so I
    can’t comment on that. However, it's worth noting that Timnit's work underscores the
    importance of interpretability in machine learning, particularly when
    it comes to the implicit bias of machine learning models trained
    on large datasets. Our work suggests that the cost of interpretability is small, even
    negligible, for many real-world applications.
  \end{quote}
\item Connection to text summarization
  \begin{quote}
    While our work is foundational, and not specific to, say natural
    language processing, the method we analyze in this paper, namely
    Determinantal Point Processes, has been used in a number of
    concrete tasks such as document summarization and video
    summarization, and has proven empirically effective. Our
    theoretical analysis provides a better understanding of these
    empirical results.
  \end{quote}
  \end{itemize}
\vspace{2mm}
